Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-05-24 paper_jasa.tex
notation for visual test added

\graphicspath{{images/}}
\usepackage{color}
\newcommand{\blue}[1]{{\color{blue} #1}}
\newcommand{\red}[1]{{\color{red} #1}}

@@

\noindent Ideally, the visual test statistic is defined (\red{should it be constructed/plotted/drawn?}) using the grammar of graphics \citep{wilkinson:1999,hadley:2009}, consisting of type, and specification of aesthetics, necessary for complete reproducibility. The visual test statistic is compared with values $T(y_0)$ using a lineup, which is defined as:

@@

\begin{dfn}\label{dfn:test}
The \textbf{(\red{Visual}) test} is defined to be 
\begin{itemize}\itemsep-3pt
\item \textbf{Reject} $H_0$ if the observer selects the observed data plot, and
\item \textbf{Fail to reject} $H_0$ if the observer selects a null plot.
\end{itemize}
\end{dfn}

@@

\noindent

Analysts typically have a choice of plots to make, and a myriad of possible options such as reference grids, for any particular purpose. This is akin to different choices of statistics in classical hypothesis testing, for example, mean, median, or trimmed mean. One is typically better than another. For two different visual test statistics of the same observed data, the one is considered to be better, if $T(y)$ is more easily distinguishable to the observer. Power is typically used to measure this characteristic of a test. Power is the probability of rejecting $H_0$ when it is not true.  We can assess power therefore both empirically based on experimental data and through theoretical considerations.

@@

\begin{dfn} \label{dfn:power}
For a lineup of $m$ plots the \textbf{power} of a visual test $T$ is defined as the probability to reject the null hypothesis for a given parameter value $\theta$:
    \begin{equation*}
      \text{Power}_T(\theta)= 
        \begin{cases} 
              \text{Type I error}=\frac1m & \text{if $\theta \in \Theta_0$,} \\
              Pr(\text{Reject } H_0) &\text{if $\theta \in \Theta^c_0$.}
        \end{cases}
    \end{equation*}
\end{dfn}

For any single observer, the power of a visual test has a lower limit of $\frac1m$ since the probability that an observer will randomly pick the test statistic under $H_0$ is $\frac1m$.

When $K$ individuals evaluate a lineup plot independently, the number of successful evaluations, $U \sim \text{Binom} (K,1/m)$  under $H_0$. This leads us to obtain the probability 
\begin{equation}\label{eqn:pvalue}
Pr(U \ge u)= \sum_{k \ge u}^K {{K \choose k} \left(\frac{1}{m}\right)^k\left(1-\frac 1m\right)^{(K-k)}}
\end{equation}
where $u$ is the observed number of successful evaluations. We use this probability to  define the $p$-value for visual test $T_{\theta}$ as follows:
\begin{dfn} \label{dfn:pval}
The $p$-value associated with the decision of a visual test $T_{\theta}$ is given \blue{as the probability to observe a value at least as extreme as the test statistic $u$, 
where $u$ is the number of successful evaluations of the lineup, which leads us to a $p$-value of $P(U \ge u)$ as given in eqn (\ref{eqn:pvalue}). Since $u$ is a discrete value with boundaries 0 and $K$,
 we can also re-formulate the $p$-value in terms of upper and lower boundaries:
 \begin{equation*}
 \begin{array}{rcll}
 P(U \ge u+1) \le &p\text{-value}& (\le 1), & \text{if }u < K, \text{and in particular for } u=0\\
( 0 \le) &p\text{-value}& \le P(U \ge u-1), & \text{if }u > 1, \text{and in particular for } u=K.\\ 
 \end{array}
 \end{equation*}
 }
\red{HH: Tried to re-write it to answer the question below.
Why is the $p$ value not just $Pr(U \ge u)$ ? (We need to modify this a bit with respect to the rejection region $R$, but there should not be a difference in terms of whether we reject or accept. }

>>>>

\graphicspath{{images/}}
\usepackage{color}
\newcommand{\blue}[1]{{\color{blue} #1}} %MM
\newcommand{\red}[1]{{\color{red} #1}} %HH
\newcommand{\green}[1]{{\color{green} #1}} %DC

@@

\noindent Ideally, the visual test statistic is defined and constructed using the grammar of graphics \citep{wilkinson:1999,hadley:2009}, consisting of type, and specification of aesthetics, necessary for complete reproducibility. The visual test statistic is compared with values $T(y_0)$ using a lineup, which is defined as:

@@

\begin{dfn}\label{dfn:test}
The \textbf{visual test}, $V_{\theta}$, is defined to be 
\begin{itemize}\itemsep-3pt
\item \textbf{Reject} $H_0$ if the observer selects the observed data plot in the lineup of $m$ plots, and
\item \textbf{Fail to reject} $H_0$ if the observer selects a null plot.
\end{itemize}
\end{dfn}

@@

\green{Ok, I'm really getting confused now. I think that it might be better to define p-value, and significance level next, and then talk about power. Power and p-value are confounded in the discussion below.}

\noindent 
Analysts typically have a choice of plots to make, and a myriad of possible options such as reference grids, for any particular purpose. This is akin to different choices of statistics in classical hypothesis testing, for example, mean, median, or trimmed mean. One is typically better than another. For two different visual test statistics of the same observed data, the one is considered to be better, if $T(y)$ is more easily distinguishable to the observer. Power is typically used to measure this characteristic of a test. Power is the probability of rejecting $H_0$ when it is not true.  We can assess power therefore both empirically based on experimental data and through theoretical considerations.

@@

\begin{dfn} \label{dfn:power}
The \textbf{power} of a visual test $V_{\theta}$ is defined as the probability to reject the null hypothesis for a given parameter value $\theta$:
    \begin{equation*}
      \text{Power}_V(\theta)= 
        \begin{cases} 
              \frac1m & \text{if $\theta \in \Theta_0$,} \\
              Pr(\text{Reject } H_0) &\text{if $\theta \in \Theta^c_0$.}
        \end{cases}
    \end{equation*}
\end{dfn}

\noindent For any single observer, the power of a visual test has a lower limit of $\frac1m$ since the probability that an observer will randomly pick the test statistic under $H_0$ is $\frac1m$. When $K$ individuals evaluate a lineup plot independently, the number of successful evaluations, $U \sim \text{Binom} (K,1/m)$  under $H_0$. This leads us to obtain the probability 
\begin{equation}\label{eqn:pvalue}
Pr(U \ge u)= \sum_{k \ge u}^K {{K \choose k} \left(\frac{1}{m}\right)^k\left(1-\frac 1m\right)^{(K-k)}}
\end{equation}
where $u$ is the observed number of successful evaluations. We use this probability to  define the $p$-value for visual test $V_{\theta}$ as follows:
\begin{dfn} \label{dfn:pval}
The $p$-value associated with the decision of a visual test $V_{\theta}$ is given \blue{as the probability to observe a value at least as extreme as the test statistic $u$, 
where $u$ is the number of successful evaluations of the lineup, which leads us to a $p$-value of $P(U \ge u)$ as given in eqn (\ref{eqn:pvalue}). Since $u$ is a discrete value with boundaries 0 and $K$,
 we can also re-formulate the $p$-value in terms of upper and lower boundaries:
 \begin{equation*}
 \begin{array}{rcll}
 P(U \ge u+1) \le &p\text{-value}& (\le 1), & \text{if }u < K, \text{and in particular for } u=0\\
( 0 \le) &p\text{-value}& \le P(U \ge u-1), & \text{if }u > 1, \text{and in particular for } u=K.\\ 
 \end{array}
 \end{equation*}
 }
\red{HH: Tried to re-write it to answer the question below.
Why is the $p$ value not just $Pr(U \ge u)$ ? (We need to modify this a bit with respect to the rejection region $R$, but there should not be a difference in terms of whether we reject or accept. }
